---
title: "DendroNetwork: a R-package to create dendrochronological provenance networks"
authors: 
  - Name: Ronald Visser
  - orcid: 0000-0001-6966-1729
  - affiliation: 1
affiliations:
   - name: Saxion University of Applied Sciences
   - index: 1
date: "`r Sys.Date()`"
bibliography: paper.bib
output: 
  html_document:
      keep_md: yes
  pdf_document: default
tags:
- dendrochronology
- network analysis
- R
- cytoscape
- tree ring
---


# Summary

(Begin your paper with a summary of the high-level functionality of your software for a non-specialist reader. Avoid jargon in this section) check: <https://joss.readthedocs.io/en/latest/submitting.html>

# Introduction

Dendrochronological analyses 

Test [@visser2021; @visser2022].

dendrochronology -> similarity relations -> network, patterns visualisation -> provenance, also wood use (papers jcaa and ijwc)


# Statement of need

There are several packages available in R for dendrochronological research [@guiterman2020; @jevsenak2018; @campelo2012; @bunn2008; @shi2019; @vandermaaten-theunissen2015; @reynolds2021; @altman2014; @rademacher; @campelo2019]. An overview with links can be found here (<https://github.com/RonaldVisser/Dendro_R>). All these packages all fill different needs, but the nice thing is that these are all interconnected in some way in depending on each other, or that they build further into different avenues. Apart from depending on dendrochronological packages, various packages from the Tidyverse [@wickham2019] are often also needed. The various relations can easily be visualised using a network, with the edges based on the dependency of two packages on each other, and and arrow indicating the direction of the dependency.

```{r creating a network of depending packages in R}
g_dendro_packages <- igraph::graph( c("dplR", "dfoliatR", "tidyverse", "dfoliatR",
                     "dplR", "dendroTools", "tidyverse", "dendroTools",
                     "dplR", "detrendeR", "tidyverse", "dplR",
                     "dplR", "fellingdateR", "tidyverse", "fellingdateR",
                     "dplR", "measuRing", "measuRing", "MtreeRing",
                     "dplR", "MtreeRing", "tidyverse", "MtreeRing",
                     "dplR", "pointRes", "tidyverse", "pointRes",
                     "dplR", "RingdateR", "tidyverse", "RingdateR",
                     "dplR", "wiad", "tidyverse", "wiad",
                     "dplR", "TRADER", "dplR", "xRing", "detrendeR", "xRing"))
plot(g_dendro_packages, edge.arrow.size=.5, vertex.color="blue", vertex.size=15, vertex.frame.color="gray", vertex.label.color="black", vertex.label.cex=0.8, vertex.label.dist=2)
```

The network shows that nearly all packages depend on dplR [@bunn2008]. The newly created package DendroNetwork fits in this ecosystem of depending packages, since it is depended on both dplR and the Tidyverse. However, it also adds a whole new world by adding network analyses through igraph [@csardi2006] to the ecosystem of dendrochronological packages. The igraph library has close connections to the tidyverse, creating a full circle and filling a hole. 

```{r creating a network of the dependencies the DendroNetwork packages}
g_DendroNetwork <- igraph::graph( c("dplR", "DendroNetwork", "tidyverse", "DendroNetwork", "igraph", "DendroNetwork", "igraph", "tidyverse", "tidyverse", "igraph"))
plot(g_DendroNetwork, edge.arrow.size=.5, vertex.color="blue", vertex.size=15, vertex.frame.color="gray", vertex.label.color="black", vertex.label.cex=0.8, vertex.label.dist=2)
plot(g_DendroNetwork+g_dendro_packages, edge.arrow.size=.5, vertex.color="blue", vertex.size=15, vertex.frame.color="gray", vertex.label.color="black", vertex.label.cex=0.8, vertex.label.dist=2)
```



# Acknowledgements

# References
